Product graphics

### 2009-12-01 product-graphics.tex
Improving display section

It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, but each category of z sums to one.

@@

Based on the modelling specification of \citep{wilkinson:1973} we define an algebra for describing these tables.

\begin{itemize}
  \item \verb|~ a| represents $f(a)$,
  \item \verb|~ a + b| represents $f(a, b)$
  \item \verb!~ a | b! represents $f(a | b)$ which is equivalent to $f(a, b)$ but with the constraint that $\forall b \sum_a f(a | b) = 1$.
\end{itemize}

@@

Once we have made the decision to map area to proportion, then we need to impose some constraints on the set of possible tilings of the plane.  Some constraints are implied by our perceptual capabilities and others by simple geometry. (In general, we could also think of partitions of higher-dimensional spaces (e.g. 3d or 4d), but we will need to project these down to two for viewing anyway, so we might as well work directly in the final space.)

\begin{itemize}
  \item Area should be proportional to weight.  Imposed by problem statement.

  \item Rectangular partitions. It is easier to compare the areas of simple shape (i.e. rectangle vs polygon), and even better when one the length of one border is fixed, so we only need compare the length of the other \citep{cleveland:1984}.  Rectangular partitions are also recursive - a rectangular partition of a rectangular area produces rectangles.  There are a limited number of shapes that do this - triangles yes, circles, no.

  \item Containment / Non-overlapping.  To be able to see the complete area of each rectangle they must be non-overlapping, and each split must be nested within the shapes of the previous level.

@@

Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$.  $a(i) \propto f(i)$ or $a(i) = k \dot a(i)$, and $a(i) = w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively.  Most of the the partitions described below use additional constraints: $h(i) = \sqrt{k} f(i)^x$ and $w(i) = \sqrt{k} f(i)^y$ where $x + y = 1$.  This ensures that it is only necessary to inspect one border to compare magnitudes.

The other chief difference between different tiling methods is whether or not they are space filling.  Space filling methods will ensure that areas are larger, but typically make it harder to compare magnitudes because scales no longer align.

The following two sections discuss particular partitions of the plane based on either a 1d or 2d pmf.  Displays of higher dimensional pmfs must be made by recursive combinations of these.

\subsection{1d primitives}

There are three 1d primitives, as displayed in Figure~\ref{fig:part-1d}:

\begin{itemize}
  \item {\bf bar}: height is proportional to value, width equally divides space. Bars allow comparison between absolute numbers, but are not space filling.  They occupy sum(values / max(values)) of the total area.  Bars can be arranged horizontally (``hbar'') or vertically (``vbar'').

  \item {\bf spine}: width is proportional to value, height equally divides space. Splines are space filling, but make it more difficult to compare absolute counts.  Splines can be arranged horizontally (``hspine''),  vertically (``vspine''), or can automatically pick their orientation (``spine'').

  \item {\bf treemap}: no restrictions on height and width. Comparing the areas is the most difficult perceptually, and are most useful when there are a large number of categories.  There are a many variations of the treemap; this paper uses the squarified treemap \citep{bruls:1999}.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.33\linewidth]{part-hbar}%
    \includegraphics[width=0.33\linewidth]{part-hspine}
    % \includegraphics[width=0.25\linewidth]{part-treemap}

  \caption{1d partitions showing proportions 0.5, 0.2, 0.2 and 0.1. From left to right: horizontal bars, horizontal spines, and a treemap.}
  \label{fig:part-1d}
\end{figure}

\subsection{2d primitives}

\begin{itemize}
  \item {\bf fluct}: width and height proportional to square root of value. Flucts allow comparisons of absolute numbers in two directions.

One special 2d case is the equal bin size plot - which all 1d x 1d primitives and the fluct collapse to when proportions are equal, as when conditioned.

Figure~\ref{fig:fluct} shows these two behaviours, looking at the relationship between happiness, health and financial status. On the left, a display of the raw proportion shows that most people are in good health and average financial standing, but it is difficult to see how happiness varies because the size of the squares is variable. On the right, conditioning on financial status and health makes all squares the same size and makes it easier to see the conditional distribution of happiness - people tend to be happier the healthier and the richer they are

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\linewidth]{part-fluct}%
    \includegraphics[width=0.5\linewidth]{part-fluct-cond}
  \caption{caption}
  \label{fig:fluct}
\end{figure}

\subsection{Plot templates}

@@

  \item {\bf Equal bin size} plot (3d): fluct and vspine, provided the final two variables are uniform.

  \item {\bf Mosaic} plot (nd).  Alternating hspines and vspines.  \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
  \item {\bf Double-decker} plot (nd).  $n-1$ hspines and 1 vspine. \citep{hofmann:2001}
  \item {\bf Treemap} (nd): n spines. \citep{shneiderman:1992}
  \item {\bf Squarified treemap} (nd): n squares. \citep{bruls:1999}

@@

Trellis graphics are another type of display that uses categorical variables to create small multiples for different subsets of the data. Visually, trellis graphics look like equal-bin-size plots with another plot drawn inside each bin.

@@

@@

It can be useful to violate the constraint that area should be proportional to weight to distinguish between zeros and very small values.  A zero weight should have zero area, but giving it positive area can be useful so it's actually possible to see!  Similarly for very small values which would otherwise occupy less than a pixel, useful to constain to minimal perceptible size.

@@

>>>>

It can be useful to display partial decompositions, i.e. just $f(x, y | z)$, which is still a 3 variable pmf, and is defined as

\[ f(x, y | z) = \frac{f(x, y, z)}{\sum_{x = a, y = b} f(a, b, z)} \]

so that each category of z sums to one, and for any fixed value of z, $f(x, y | z)$ is a valid pmf.

\[ \forall b \sum_a f(a | b) = 1 \]

@@

Based on the modelling specification of \citet{wilkinson:1973} we define an algebra for describing these tables.

\begin{itemize}
  \item \verb|~ a| represents $f(a)$,
  \item \verb|~ a + b| represents $f(a, b)$
  \item \verb!~ a | b! represents $f(a | b)$
\end{itemize}

@@

Once we have made the decision to map area to proportion, we need to impose some constraints on the set of possible tilings of the plane. In general, we could also think of partitions of higher-dimensional spaces (e.g. 3d or 4d), but we will need to project these down to two for viewing anyway, so there is little disadvantage to working in 2d.

Some constraints are implied by our perceptual capabilities and others by geometry. On top of the constraint that area must be proportional to value, we add two additional constraints:

\begin{itemize}

  \item Partitions must be rectangular. It is easier to compare the areas of simple shape (i.e. rectangle vs polygon), and even better when one the length of one border is fixed, so we only need compare the length of the other \citep{cleveland:1984}. Rectangular partitions are also recursive in the sense that we can always tile a rectangle with smaller rectangles. This is not true for most shapes.

  \item Partitions must be disjoint. To be able to see the complete area of each rectangle they must be non-overlapping, and each split must be nested within the shapes of the previous level. Note that we do not impose the constraint that the tiling be space-filling. Space filling methods use will ensure that areas are larger, but make it harder to compare magnitudes because scales no longer align.

@@

Let $a(i)$ give the area assigned to category $i$, which has proportion $f(i)$. Since we are using rectangles, $a(i) = k \dot a(i) = k \dot w(i) \times h(i)$ where $w$ and $h$ represent width and height respectively. Certain partitions also satisfy the constraint that $h(i) = \sqrt{k} f(i)^x$ and $w(i) = \sqrt{k} f(i)^y$ where $x + y = 1$. This is useful because it is only necessary to inspect one border to compare magnitudes.

The following two sections describe the partitions currently used to produce area graphics. Section~\ref{sub:part-1d} discusses partitions for 1d pmfs: bars, spines and tiles, Section~\ref{sub:part-2d} discusses one partition for 2d pmfs and Section~\ref{sub:part-nd} describes how the 1d and 2d primitives can be recursively combined to generate a partitioning for any higher dimension pmf and discusses how existing area graphics fall into this framework.

\subsection{1d primitives}
\label{sub:part-1d}

There are three 1d primitives, as shown in Figure~\ref{fig:part-1d}:

\begin{itemize}
  \item {\bf bar}s: height is proportional to value, width equally divides space. Bars allow comparison between absolute numbers, but are not space filling.  They occupy sum(values / max(values)) of the total area.  Bars can be arranged horizontally (``hbar'') or vertically (``vbar'').  Not space filling.

  \item {\bf spine}s: width is proportional to value, height equally divides space. Splines are space filling, but make it more difficult to compare absolute counts.  Splines can be arranged horizontally (``hspine''),  vertically (``vspine''), or can automatically pick their orientation (``spine'').  Space filling.

  \item {\bf tile}s: tile the plane with rectangles, trying to keep the aspect ratio of each rectangle close to 1. Comparing the areas is the most difficult perceptually, and are most useful when there are a large number of categories.  The is partitioning used by the squarified treemap \citep{bruls:1999}.  Not space filling.

@@

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.33\linewidth]{part-hbar}%
    \includegraphics[width=0.33\linewidth]{part-hspine}%
    \includegraphics[width=0.33\linewidth]{part-treemap}

  \caption{1d partitions showing the distribution of happiness.  From left to right: bars, spines and tiles.}
  \label{fig:part-1d}
\end{figure}

The appearance of bars and splines is identical for uniformly distributed data.

\subsection{2d primitives}
\label{sub:part-2d}

We are currently aware of one 2d primitive, the {\bf fluct}: where width and height are proportional to the square root of the proportion. Flucts allow comparisons of relative magnitudes numbers in two directions. Each rectangle is arranged on a regular grid formed by the levels of the two variables, and thus the fluct is not space filling. A fluct displays a joint 2d distribution, unlike the combination of any two 1d partitions which display a conditional distribution.

One special case of the fluct is the equal bin size plot which occurs when the two variables are jointly uniformly distributed, usually as a result of conditioning. The equal bin size plot is useful to visualise missing combinations, or when supplemented with a third variable, to visualise the distribution conditional on two other variables.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\linewidth]{part-fluct}%
    \includegraphics[width=0.5\linewidth]{part-fluct-cond}
  \caption{Distribution of age (in decades) and marital status.  All combinations of the two variables are present.}
  \label{fig:fluct}
\end{figure}

\subsection{Combinations}
\label{sub:part-nd}

To visualise higher-dimensional pmfs, we partition recursively using the primitives described above.  That is, to display $f(x, y, z)$ we first create a partition based on $f(z)$ then partition each of those pieces based on $f(y | z)$ and again by $f(x | y, z)$.  Figure \ref{fig:recursive} illustrates this process.

Figure~\ref{fig:fluct} uses a fluct and a vspine to explore the relationship between happiness, health and financial status. On the left, a fluctuation diagram shows that most people are in good health and average financial standing, but it is difficult to see how happiness varies because the size of the squares is variable. Conditioning on financial status and health produces the equal bin size plot on the right and makes it easier to see the conditional distribution of happiness - people tend to be happier the healthier and richer they are.

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.5\linewidth]{part-nd}%
    \includegraphics[width=0.5\linewidth]{part-nd-1}
  \caption{caption}
  \label{fig:part-nd}
\end{figure}

@@

  \item {\bf Equal bin size} plot (3d): fluct and vspine, provided the first two variables are uniform.

  \item {\bf Mosaic} plot (nd).  Alternating hspines and vspines.  \citep{hartigan:1984,hartigan:1981,friendly:1994,hofmann:2003}
  \item {\bf Double-decker} plot (nd).  $n-1$ hspines and 1 vspine. \citep{hofmann:2001}
  \item {\bf Treemap} (nd): n spines. \citep{shneiderman:1992}
  \item {\bf Squarified treemap} (nd): n tiles. \citep{bruls:1999}

@@

Trellis graphics \citep{becker:1996}, as known as lattice or facetted or
conditioned graphics are another related display. They uses categorical
variables to generate multiple panels, each containing a plot of the subset of
the data. Trellised area plots also fall into our framework and can be
created by conditioning on the trellising variables.

@@

There are a number of cases in which it is useful to relax the constraints that we defined above.  These are described in sequence below.

@@

It can be useful to violate the constraint that area should be proportional to weight to distinguish between zeros and very small values.  A zero weight should have zero area, but giving it positive area can be useful so it's actually possible to see!  Similarly for very small values which would otherwise occupy less than a pixel, useful to constrain to minimal perceptible size.

@@

\subsection{Missing values}
\label{sub:missing_values}
